---
title: "Ottawa eBird Checklist Data Analysis"
author: "Andrew Habrich"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    highlight: tango
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
## HEADER---------------------------
## Script name: X2-singlecity_figures.Rmd
## 
## Purpose of script: This script will generate an RMD html file with some preliminary figures and results
## 
## Author: Andrew Habrich
## 
## Date Created: 2025-01-30
## Date last Modified: 2025-03-17
## 
## Email:
## - [andrhabr@gmail.com](mailto:andrhabr@gmail.com){.email}
## - [andrewhabrich@cmail.carleton.ca](mailto:andrewhabrich@cmail.carleton.ca){.email}
```

```{r knitsetup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 4)
```

```{r packagesetup, message=FALSE, warning=FALSE, include=FALSE}
pacman::p_load(sf, terra, tidyterra, rnaturalearth, mapview, mgcv, gam,
lme4, mgcv, modelsummary, tidyverse, ggpubr, RColorBrewer, easystats, tictoc)
```

```{r chkldata, include=FALSE}
chkl_bufferlist <- readRDS("./_unittest/derived_output/Ottawa_chkl_bufferlist.rds")
```

All figures for just Ottawa-Gatineau eBird data from 2010-2022

```{r correlations, eval=FALSE, fig.fullwidth=T, include=FALSE, paged.print=TRUE}
# Plot the correlation between the landscape metrics and raster data for each of the different buffer radius dataframes
chkl_buffer_correlations <- list() 
for (i in 1:length(chkl_bufferlist))
{ 
chklcor <- chkl_bufferlist[[1]] %>% select(specrich, boundarydist_m,
uacentrdist_m, ai, iji, contag, mesh, starts_with("p_"), starts_with("bv_"), starts_with("enn")) %>% 
    correlation(redundant = T)
  
chklcorplot <- as.matrix(chklcor) %>% 
  plot(title = NULL) + 
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) 

chkl_buffer_correlations[[i]] <- chklcorplot 
rm(chklcor,chklcorplot) 
}

#chkl_buffer_correlations 
# ggarrange(plotlist = chkl_buffer_correlations, ncol = 2, nrow = 2, align = "hv", 
#           common.legend = T, legend = "right",
# labels = names(chkl_bufferlist))
print(chkl_buffer_correlations)
```

# Histograms

## Species richness distribution in each buffer

```{r spprich_hist, echo=FALSE, message=FALSE, warning=FALSE}
full_histogram <- chkl_bufferlist[[1]] %>% 
  gghistogram(x = "specrich", binwidth = 1, add = "mean", 
              rug = T, add_density = T, 
              alpha = 0.25, color = "black", fill = "darkgreen", 
              title = "Distribution of Spp.Rich. from eBird Checklists in Ottawa")
full_histogram
# filt_histogram <- chkl_bufferlist[[1]] %>% 
#   filter(p_greencover >= 25 & p_greencover <= 75) %>%
#   gghistogram(x = "specrich", binwidth = 1, add = "mean", 
#               rug = T, add_density = T, 
#               alpha = 0.25, color = "black", fill = "forestgreen", 
#               title = "Distribution of Spp.Rich. from eBird Checklists in Ottawa") 
# filt_histogram
```

## Distribution of proportion tree values in each buffer

```{r ptreehist, echo=FALSE, message=FALSE, warning=FALSE}
chkl_hist_treed <- list() 
for (i in 1:length(chkl_bufferlist)) { 
  histogram <- chkl_bufferlist[[i]] %>% 
  gghistogram(x = "p_tree", binwidth = 2, add = "mean", 
              rug = T, add_density = T, 
              alpha = 0.25, color = "black", fill = "green", title = names(chkl_bufferlist[i])) 
chkl_hist_treed[[i]] <- histogram 
rm(histogram) 
} 
ggarrange(plotlist = chkl_hist_treed, ncol = 2, nrow = 2, common.legend = T, legend = "right")
```

## Distribution of combined 'green' LC type in each buffer

```{r pgreen_hist, echo=FALSE, message=FALSE, warning=FALSE}
## NOTE THIS IS THE COMBINED TREE, SHRUB, GRASS LC types
chkl_hist_green <- list() 
for (i in 1:length(chkl_bufferlist)) { 
  histogram <- chkl_bufferlist[[i]] %>% 
    gghistogram(x = "p_greencover", binwidth = 2, add = "mean", 
                rug = T, add_density = T,
                alpha = 0.25, color = "forestgreen", fill = "grey",
                title = names(chkl_bufferlist[i])) 
  chkl_hist_green [[i]] <- histogram 
  rm(histogram) 
} 
histplot <- ggarrange(plotlist = chkl_hist_green, ncol = 2, nrow = 2, common.legend = T, legend = "right") 
histplot

# export as figure for later
histplot %>% 
  ggexport(filename = "./_unittest/derived_output/Ott_chkl_greencover.png")
```

## Greencover values bound between 25% and 75%

```{r pgreen_selectedhist, echo=FALSE, warning=FALSE}
chkl_hist_greenfiltered <- list() 
for (i in 1:length(chkl_bufferlist)) { 
  histogram <- chkl_bufferlist[[i]] %>% 
    filter(p_greencover >= 25 & p_greencover <= 75) %>% 
    gghistogram(x = "p_greencover", binwidth = 1, add = "mean", 
                rug = T, add_density = T, 
                alpha = 0.25, color = "black", fill = "forestgreen", 
                title = names(chkl_bufferlist[i]))
  chkl_hist_greenfiltered [[i]] <- histogram 
  rm(histogram) 
}
ggarrange(plotlist = chkl_hist_greenfiltered, ncol = 2, nrow = 2, common.legend = T, legend = "right")
```

## Plot the distribution of proportion builtup values

```{r pbuilt_hist, echo=FALSE, warning=FALSE}
chkl_hist_builtup <- list() 
for (i in 1:length(chkl_bufferlist)) { 
  histogram <- chkl_bufferlist[[i]] %>% 
    filter(p_greencover >= 25 & p_greencover <= 75) %>% 
    gghistogram(x = "p_builtup", binwidth = 2, add = "mean", 
                rug = T, add_density = T, 
                alpha = 0.25, color = "black", fill = "grey", 
                title = names(chkl_bufferlist[i]))
  chkl_hist_builtup[[i]] <- histogram 
  rm(histogram) 
  } 
ggarrange(plotlist = chkl_hist_builtup, ncol = 2, nrow = 2, common.legend = T, legend = "right")
```

## Distribution of TOTAL building volumes in buffer

```{r bvsum_hist, echo=FALSE, warning=FALSE}
chkl_hist_bv <- list()
for (i in 1:length(chkl_bufferlist)) { 
  histogram <- chkl_bufferlist[[i]] %>% 
    filter(p_greencover >= 25 & p_greencover <= 75) %>% 
    mutate(bv_sum_scale = bv_sum/1000) %>% 
    gghistogram(x = "bv_sum_scale", add = "mean", 
                rug = T, add_density = T, 
                alpha = 0.25, color = "black", fill = "violet",
                title = names(chkl_bufferlist[i])) 
  chkl_hist_bv[[i]] <- histogram 
  rm(histogram) 
  } 
ggarrange(plotlist = chkl_hist_bv, ncol = 2, nrow = 2, 
          common.legend = T, legend = "right")
```

## Distribution of MEAN building volumes in buffer

```{r bvmean_hist, echo=FALSE, warning=FALSE}
chkl_hist_bv <- list()
for (i in 1:length(chkl_bufferlist)) { 
  histogram <- chkl_bufferlist[[i]] %>% 
    filter(p_greencover >= 25 & p_greencover <= 75) %>% 
    gghistogram(x = "bv_mean", add = "mean", 
                rug = T, add_density = T, 
                alpha = 0.25, color = "black", fill = "purple",
                title = names(chkl_bufferlist[i])) 
  chkl_hist_bv[[i]] <- histogram 
  rm(histogram) 
  } 
ggarrange(plotlist = chkl_hist_bv, ncol = 2, nrow = 2, 
          common.legend = T, legend = "right")
```

# Scatterplots

## Scatterplot of the greencover and built volume

```{r scatter_green_bv, echo=FALSE, message=FALSE, warning=FALSE}
chkl_plot_forest.bv <- list()
for (i in 1:length(chkl_bufferlist)) { 
  scatterplot <- chkl_bufferlist[[i]] %>% 
    filter(p_greencover >= 25 & p_greencover <= 75) %>% 
    ggscatter(x = "p_greencover", y = "bv_mean", 
              alpha = 0.25, title = names(chkl_bufferlist[i]),
              color = "forestgreen") 
  chkl_plot_forest.bv[[i]] <- scatterplot 
  rm(scatterplot) 
  } 

ggarrange(plotlist = chkl_plot_forest.bv, ncol = 2, nrow = 2, 
          common.legend = T, legend = "right")
```

## Scatterplot of the builtup and built volume

```{r scatter_pbuilt_bv, echo=FALSE, message=FALSE, warning=FALSE}
chkl_plot_built.bv <- list()
for (i in 1:length(chkl_bufferlist)) { 
  scatterplot <- chkl_bufferlist[[i]] %>% 
    filter(p_greencover >= 25 & p_greencover <= 75) %>% 
    ggscatter(x = "p_builtup", y = "bv_mean", 
              alpha = 0.5, title = names(chkl_bufferlist[i])) 
  chkl_plot_built.bv[[i]] <- scatterplot 
  rm(scatterplot) 
} 

ggarrange(plotlist = chkl_plot_built.bv, ncol = 2, nrow = 2, 
          common.legend = T, legend = "right")
```

## Scatterplot between both measures of checklist distance

```{r chkldist, echo=FALSE, message=FALSE, warning=FALSE}
chkl_distances <- chkl_bufferlist[[1]] %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  filter(boundarydist_m > 1000) %>% 
  ggscatter(x = "boundarydist_m", y = "uacentrdist_m", 
            alpha = 0.5, 
            title = "Checklist distance to urban area boundary vs centroid", 
            color = "grey", fill = "grey", 
            add = "reg.line", add.params = list(color = "red"), 
            cor.coef = T, cor.method = "pearson") 
print(chkl_distances)

## export a figure 
ggexport(chkl_distances, 
         filename = "./_unittest/derived_output/Ott_chkl_distances.png")
```

## Check the scatterplot of checklist distances and the p_greencover

```{r chkldists_plots, echo=FALSE, fig.fullwidth=T, message=FALSE, warning=FALSE}
### Distance to urban area boundary
chkl_plot_bounddist.greencover <- list()
for (i in 1:length(chkl_bufferlist)) { 
  scatterplot <- chkl_bufferlist[[i]] %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  filter(boundarydist_m > 1000) %>% 
    ggscatter(y = "boundarydist_m", x = "p_greencover", 
              alpha = 0.5, title = names(chkl_bufferlist[i]),
              color = "olivedrab", fill = "black") 
  chkl_plot_bounddist.greencover[[i]] <- scatterplot 
  rm(scatterplot) 
  } 
bd <- ggarrange(plotlist = chkl_plot_bounddist.greencover, 
          ncol = 2, nrow = 2, common.legend = T, legend = "right") 
bd
### Distance to urban area centroid 
chkl_plot_centrdist.greencover <- list() 
for (i in 1:length(chkl_bufferlist)) { 
  scatterplot <- chkl_bufferlist[[i]] %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  filter(boundarydist_m > 1000) %>% 
  ggscatter(y = "uacentrdist_m", x = "p_greencover", 
           alpha = 0.5, title = names(chkl_bufferlist[i]),
           color = "grey", fill = "black") 
  chkl_plot_centrdist.greencover[[i]] <- scatterplot 
  rm(scatterplot) 
  } 
ud <- ggarrange(plotlist = chkl_plot_centrdist.greencover, 
          ncol = 2, nrow = 2, common.legend = T, legend = "right")
ud

#ggarrange(plotlist = list(bd,ud), nrow = 2, ncol = 1, 
#          labels = c("Dist. to urban boundary",
#                     "Dist. to urban centroid"))
```

# Species richness as the response variable

## Chkl distances against species richness

```{r chkldists_sprich_plots, echo=FALSE, fig.fullwidth=T, message=FALSE, warning=FALSE}
### Distance to urban area boundary
chkl_plot_bounddist.spr <- list()
for (i in 1:length(chkl_bufferlist)) { 
  scatterplot <- chkl_bufferlist[[i]] %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  filter(boundarydist_m > 1000) %>% 
    ggscatter(x = "boundarydist_m", y = "specrich", 
              alpha = 0.5, title = names(chkl_bufferlist[i]),
              color = "olivedrab", fill = "black",
              add = "reg.line", add.params = list(color = "red")) 
  chkl_plot_bounddist.greencover[[i]] <- scatterplot 
  rm(scatterplot) 
  } 
bd <- ggarrange(plotlist = chkl_plot_bounddist.greencover, 
          ncol = 2, nrow = 2, common.legend = T, legend = "right") 
bd
### Distance to urban area centroid 
chkl_plot_centrdist.spr <- list() 
for (i in 1:length(chkl_bufferlist)) { 
  scatterplot <- chkl_bufferlist[[i]] %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  filter(boundarydist_m > 1000) %>% 
  ggscatter(x = "uacentrdist_m", y = "specrich", 
           alpha = 0.5, title = names(chkl_bufferlist[i]),
           color = "grey", fill = "black",
           add = "reg.line", add.params = list(color = "red")) 
  chkl_plot_centrdist.greencover[[i]] <- scatterplot 
  rm(scatterplot) 
  } 
ud <- ggarrange(plotlist = chkl_plot_centrdist.greencover, 
          ncol = 2, nrow = 2, common.legend = T, legend = "right")
ud

#ggarrange(plotlist = list(bd,ud), nrow = 2, ncol = 1, 
#          labels = c("Dist. to urban boundary",
#                     "Dist. to urban centroid"))
```

## Species richness vs prop. greencover

```{r spprich_scatter, echo=FALSE, fig.fullwidth=T, message=FALSE, warning=FALSE}
chkl_plot_specrich <- list() 
for (i in 1:length(chkl_bufferlist)) { 
  scatterplot <- chkl_bufferlist[[i]] %>% 
    filter(p_greencover >= 25 & p_greencover <= 75) %>% 
    filter(boundarydist_m > 1000) %>% 
    ggscatter(x = "p_greencover", y = "specrich", 
              alpha = 0.10, title = names(chkl_bufferlist[i]), 
              color = "forestgreen", fill = "black", 
              add = "reg.line", add.params = list(color = "red"), 
              cor.coef = T, cor.method = "pearson")
  chkl_plot_specrich[[i]] <- scatterplot 
  rm(scatterplot) 
 } 
ggarrange(plotlist = chkl_plot_specrich, 
          ncol = 2, nrow = 2, common.legend = T, legend = "right")
```

## Species richness vs ai, iji, mesh, contag, ENN

For the 500m buffer radius

```{r spprich_lsmetrics, echo=FALSE, fig.fullwidth=T, message=FALSE, warning=FALSE}
df500m <- chkl_bufferlist[[1]]
ai <- df500m %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  ggscatter(x = "ai", y = "specrich", alpha = 0.5, 
            title = "AI", color = "forestgreen", fill = "green", 
            add = "reg.line", add.params = list(color = "black")) 
iji <- df500m %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  ggscatter(x = "iji", y = "specrich", alpha = 0.5, 
            title = "IJI", color = "forestgreen", fill = "green", 
            add = "reg.line", add.params = list(color = "black")) 
contag <- df500m %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  ggscatter(x = "contag", y = "specrich", alpha = 0.5, 
            title = "Contagion index", 
            color = "forestgreen", fill = "green", 
            add = "reg.line", add.params = list(color = "black")) 
mesh <- df500m %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  ggscatter(x = "mesh", y = "specrich", alpha = 0.5, 
            title = "MESH index", 
            color = "forestgreen", fill = "green", 
            add = "reg.line", add.params = list(color = "black"))
ennm <- df500m %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  ggscatter(x = "ennm_tree", y = "specrich", alpha = 0.5, 
            title = "Mean ENN distance", 
            color = "forestgreen", fill = "green", 
            add = "reg.line", add.params = list(color = "black"))
ennsd <- df500m %>% 
  filter(p_greencover >= 25 & p_greencover <= 75) %>% 
  ggscatter(x = "ennsd_tree", y = "specrich", alpha = 0.5, 
            title = "SD ENN distance", 
            color = "forestgreen", fill = "green", 
            add = "reg.line", add.params = list(color = "black"))

ggarrange(plotlist = list(ai, iji, contag, mesh, ennm, ennsd), nrow = 3, ncol = 2)
```

# Modelling: GLMs and GAMs

```{r modelling, message=FALSE, warning=FALSE, include=FALSE}
pacman::p_load(mgcv, lme4, MuMIn, easystats, modelsummary, DHARMa, knitr) 
### NEED TO ACCOUNT FOR SAMPLING EFFORT (distance travelled, duration)
### NEED TO ACCOUNT FOR TEMPORAL SAMPLING (month; year was a flat line)
```

## Generalized linear model

```{r simplemodel, message=FALSE, warning=FALSE, include=FALSE}
df <- chkl_bufferlist[[1]] %>% 
    filter(p_greencover > 25 & p_greencover < 75)
glm_mod <- glm(specrich ~ 
               p_greencover + bv_mean + uacentrdist_m + ennm_tree +
               ai + iji + contag + mesh + 
               offset(log(duration_minutes)) + offset(log(effort_distance_km)), 
               family = "gaussian", data = df) 
```

```{r modelstr, echo=TRUE, message=FALSE, warning=FALSE}
#Using this model
find_formula(glm_mod)$conditional
```

```{r glm_table, echo=FALSE, message=FALSE, warning=FALSE}
#Create a list and generate a model for each buffer radius
glm_mods <- list()
for (i in 1:length(chkl_bufferlist)) { 
  df <- chkl_bufferlist[[i]] %>% 
    filter(p_greencover > 25 & p_greencover < 75)
  glm_mod <- glm(specrich ~ 
                   p_greencover + bv_mean + uacentrdist_m + ennm_tree +
                   ai + iji + contag + mesh + 
                   offset(log(duration_minutes)) + offset(log(effort_distance_km)), 
                 family = "gaussian", data = df) 
glm_mods[[i]] <- glm_mod
names(glm_mods)[i] <- names(chkl_bufferlist)[i]
rm(glm_mod) 
}
# Compare the performance of the models
comparetab <- compare_performance(glm_mods, metrics = "all", estimator = "REML")
#kable table
kable(comparetab)
```

```{r modelparameters, echo=FALSE, message=FALSE, warning=FALSE}
model_res <- model_parameters(glm_mods[[1]], robust = T)
plot(model_res)
# model_param <- standardize_parameters(model = glm_mods[[1]], method = "refit")
# plot(model_param)
```

# Generalized additive model

```{r nullgams, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
gam_null1 <- gam(specrich ~ 1 + offset(log(duration_minutes)), 
                 family = "gaussian", data = chkl_bufferlist[[1]],
                 subset = p_greencover > 25 & p_greencover < 75) 

gam_null2 <- gam(specrich ~ 1 + offset(log(effort_distance_km)), 
                 family = "gaussian", data = chkl_bufferlist[[1]],
                 subset = p_greencover > 25 & p_greencover < 75) 

gam_null3 <- gam(specrich ~ 1 + 
                   offset(log(duration_minutes)) + 
                   offset(log(effort_distance_km)), 
                 family = "gaussian", data = chkl_bufferlist[[1]],
                 subset = p_greencover > 25 & p_greencover < 75)
#model_performance(gam_null1, gam_null2, gam_null3)

```

## GAMS with smoothing terms and offsets for sampling effort

```{r gams, echo=FALSE, message=FALSE, warning=FALSE}
df <- chkl_bufferlist[[1]] %>% 
  filter(p_greencover > 25 & p_greencover < 75)
gam_mods <- mgcv::gam(specrich ~ 
                  s(p_greencover, k = 12) + bv_mean + 
                  ai + iji + contag + mesh + ennm_tree +
                  month + uacentrdist_m +
                  offset(log(duration_minutes)) + 
                  offset(log(effort_distance_km)), 
                family = "gaussian", data = df, method = "REML") 
model_parameters(gam_mods)
```

```{r gam_viz, echo=FALSE, message=FALSE, warning=FALSE}
# ## check the results of the gam
#par(mfrow = c(2, 2)) 
#gam.check(gam_mods)
performance::check_model(gam_mods, residual_type = "normal")

# Plot the fitted line of the smoothed p_greencover in a ggplot
# ggplot(chkl_bufferlist[[1]],
#        aes(x = p_greencover, y = specrich)) +
#   geom_point() +
#   geom_line(colour = "red", linewidth = 1.2,
#             aes(y = fitted(gam_mods))) +
#   theme_bw()
```
